Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales

### 2020-05-23 reports/article.Rmd
Minor formatting. Fix #23.

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland, PhD`*`^1,2,3^, Erik Bulow, MSc`*`^2,4^, Erik Lenguerrand, MSc^5^, Prof Ashley Blom, PhD^5,6^, Prof J. Mark Wilkinson, PhD^7^, Adrian Sayers, MSc^5^, Prof Ola Rolfson, PhD^2,4^, Prof Nils P. Hailer, PhD^1,2^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

```{r setup, include = FALSE}
options(
  digits            = 2,
  OutDec            = "\xB7",
  knitr.kable.NA    = '',
  repos = list(CRAN = "https://cran.rstudio.com/")
)
knitr::opts_chunk$set(
  echo = FALSE,
  out.width  = "100%", 
  out.height = "100%"
)
library(tidyverse)

@@

@@

## Affiliations

## Addrasses

AG: Andersrogatan 5, SE-621 45 Visby, SWEDEN  
EB/OR: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenbutg, SWEDEN   
EL/AB/AS: Musculoskeletal Research Unit, University of Bristol, Beacon House, Queens Road, Bristol, BS8 1QU, UK  
JMW: Department of Oncology and Metabolism, The Medical School, Beech Hill Road, Sheffield, S10 2RX, UK  
NPH: Akademiska sjukhuset, 751 85 Uppsala, SWEDEN

## Correspondence 
Erik Bulow  
+ 46 70 08 234 28  
erik.bulow@vgregion.se  
Postal: Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45 Gothenburg, SWEDEN

##### PAGE BREAK

```{r}
# Change to ugly decimal point as required by LR
ud <- function(x) gsub("\\.", "\xB7", x)

shar_tot <- nrow(df)
shar_female <- df$P_Sex == "Female"
age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age)) %>% ud()
sex_shar <- 
  sprintf(
    "%.0f %% [%s of %s]", 
    mean(shar_female) * 100, 
    format(sum(shar_female), big.mark = ",", decimal.mark = "\xB7"), 
    format(shar_tot, big.mark = ",", decimal.mark = "\xB7")
) %>% ud()
deaths_shar <- 
  sprintf(
    "%.02f %% [%.0f of %s]", 
    mean(df$death90) * 100, 
    sum(df$death90), 
    format(nrow(df), big.mark = ",")
  ) %>% ud()

shar_summary <- sprintf(
  "(mean [range] age %s; %s females; %s deaths within 90 days)",
  age_shar, sex_shar, deaths_shar
) %>% ud()

extr <- function(col, reg) 
  subset(table1_combined_tot, startsWith(what, col), reg, drop = TRUE)
extr_perc <- function(...) 
  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE) %>% ud()

njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
# Same range by design (to match Swedish cohort)
njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age)) %>% ud()
njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ",")) %>% ud()
njr_deaths <- 
  sprintf(
    "%.02f %% [%.0f of %s]", 
    (N_njr$death / N_njr$tot) * 100, 
    N_njr$death, 
    format(N_njr$tot, big.mark = ",")
  ) %>% ud()
njr_summary <- sprintf(
  "(mean [range] age %s; %s females; %s deaths within 90 days)",
  njr_age, njr_sex, njr_deaths
) %>% ud()

get_auc <- function(mod) {
  subset(
    brlasso_tbl_auc, 
    Model == mod & 
      `Age as` %in% c(NA, "main effect"), 
    "AUC", 
    drop = TRUE
  ) %>% 
  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI ",., fixed = TRUE))} %>% 
  ud()
}

@@

**Background: ** Mortality within 90 days after total hip arthroplasty (THA) due to osteoarthritis is low, but the risk of short-term mortality is nevertheless essential to discuss preoperative between patients and surgeons. We aimed to develop a parsimonious prediction model with internal and external validation of discrimination and calibration, and to compare it to prediction models based on existing co-morbidity indices.

**Methods: **We performed a prospective multi-national population-based registry study. We included `r format(nrow(df), big.mark = ",")` patients with cemented THA due to osteoarthritis from the Swedish Hip Arthroplasty Register for model derivation and internal validation, as well as `r format(N_njr$tot, big.mark = ",")` patients from the National Joint Register from England and Wales for external model validation. A model was developed using a bootstrap ranking procedure with a least absolute shrinkage and selection operator (LASSO) logistic regression model combined with piecewise linear regression. Discriminative ability was evaluated by the area under the receiver operating characteristics curve (AUC). Calibration belt plots were used for model calibration.

**Finindgs: ** We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity. The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`. This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices. The model was well calibrated for predicted probabilities up to 5%.

**Interpretation: ** We developed a model of 90-day mortality after THA based on age, sex, and the presence of four defined co-morbidities. This may facilitate individualized risk assessment prior to one of the most common surgical interventions. We published a web-calculator to aid clinical decision-making (`r shiny_url`).

**Funding: ** We had no external funding.

##### PAGE BREAK

# Summary box

What is already known:

- Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality.
- Available diagnosis code-based co-morbidity instruments are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) classification is imprecise.

What this study adds:

- Our results indicate that the risk of early postoperative mortality after cemented THA can be pre-operatively assessed by a parsimonious prediction model. 
- The presented model, with its accompanying web calculator, could facilitate a shared decision-making between physicians and their patients in need of THA.

##### PAGE BREAK

@@

Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process. In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure. 90-day mortality after THA performed due to osteoarthritis is low, ranging from `r 0.2` % to `r 0.6` % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.

@@

Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation). Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above `r 0.7`, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples. We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally. An AUC with 95 % confidence interval was calculated for the model as-is. Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales. An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.

@@

`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel). `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100)) %>% ud()` of those patients died within 90 days and no one was censored before that. The unadjusted risk of 90-day mortality was therefore `r ud(surv90d)`. Further characteristics of the study population are presented in Table \@ref(tab:tab1). `r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser. The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.

In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel). Their unadjusted risk of 90 day mortality was `r ud(njr_surv)`, which was higher compared to the derivation cohort from Sweden.

@@

The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal validation `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)). Calibration of the re-calibrated model was slightly inferior compared to the internal calibration. Predicted probabilities between `r 0.5` % and `r 1.5` % were lower than observed proportions and the estimated 95 % confidence bands were wider. Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).

@@

Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE) %>% ud()`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %). We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.

@@

In addition to data validity, the indications for choosing between different fixation techniques differ in Sweden compared to England and Wales. Cementation is the most commonly used fixation principle in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants. To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden. Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below `r 1.5` % underestimated the observed proportion of deaths in the external validation cohort from the NJR.

We investigated whether the effect of age on mortality was non-linear using restricted cubic splines. A similar approach might be relevant for BMI, which we only modelled as a linear main effect. A U-shaped association between mortality and BMI categorized as underweight (below `r 18.5` kg/m^2^), normal (`r 8.5` to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel. Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

@@

##### PAGE BREAK

@@

>>>>

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

```{r setup, include = FALSE}
options(
  digits            = 2,
  knitr.kable.NA    = '',
  repos = list(CRAN = "https://cran.rstudio.com/")
)
knitr::opts_chunk$set(
  echo = FALSE,
  out.width  = "100%", 
  out.height = "100%"
)
library(tidyverse)

@@

# Affiliation and contact

@@

**Correspondence:** 
erik.bulow@vgregion.se. 
Postal: Svenska Höftprotesregistret, Registercentrum Västra Götaland, SE-413 45, Sweden

**OBJECTIVES** To develop and validate a personalized model to predict 90-day mortality after total hip arthroplasty (THA).

**DESIGN** Prospective multi-national population-based study.

**SETTING** Swedish Hip Arthroplasty Registry (SHAR) data for model derivation and internal validation, and data from the National Joint Registry (NJR) for England and Wales for external validation.

```{r}
shar_tot <- nrow(df)
shar_female <- df$P_Sex == "Female"
age_shar <- sprintf("%.0f [%.0f - %.0f]", mean(df$P_Age), min(df$P_Age), max(df$P_Age))
sex_shar <- 
  sprintf(
    "%.0f %% [%s of %s]", 
    mean(shar_female) * 100, 
    format(sum(shar_female), big.mark = ","), 
    format(shar_tot, big.mark = ",")
)
deaths_shar <- 
  sprintf(
    "%.02f %% [%.0f of %s]", 
    mean(df$death90) * 100, 
    sum(df$death90), 
    format(nrow(df), big.mark = ",")
  )

shar_summary <- sprintf(
  "(mean [range] age %s; %s females; %s deaths within 90 days)",
  age_shar, sex_shar, deaths_shar
)

extr <- function(col, reg) 
  subset(table1_combined_tot, startsWith(what, col), reg, drop = TRUE)
extr_perc <- function(...) 
  gsub("([\\d,]*)( )(\\()(\\d*)(\\.\\d*)(\\))", "\\4 \\% \\[\\1 of", extr(...), perl = TRUE)

njr_age_mean <- as.numeric(gsub("(.*) (.*)", "\\1", extr("Age", "NJR"), perl = TRUE))
# Same range by design (to match Swedish cohort)
njr_age <- sprintf("%.0f [%.0f - %.0f]", njr_age_mean, min(df$P_Age), max(df$P_Age))
njr_sex <- sprintf("%s %s]", extr_perc("Sex", "NJR"), format(N_njr$tot, big.mark = ","))
njr_deaths <- 
  sprintf(
    "%.02f %% [%.0f of %s]", 
    (N_njr$death / N_njr$tot) * 100, 
    N_njr$death, 
    format(N_njr$tot, big.mark = ",")
  )
njr_summary <- sprintf(
  "(mean [range] age %s; %s females; %s deaths within 90 days)",
  njr_age, njr_sex, njr_deaths
)
```

**PARTICIPANTS** `r format(nrow(df), big.mark = ",")` patients with cemented total hip arthroplasty (THA) due to osteoarthritis recorded in the SHAR, and `r format(N_njr$tot, big.mark = ",")` similarly selected patients recorded in the NJR.

**MAIN OUTCOME MEASURES** Predicted probability of death for any reason within 90 days after primary THA, estimated using a bootstrap ranking procedure together with a least absolute shrinkage and selection operator (LASSO) logistic regression model and evaluated by areas under the curve (AUC).

```{r}
get_auc <- function(mod) {
  subset(
    brlasso_tbl_auc, 
    Model == mod & 
      `Age as` %in% c(NA, "main effect"), 
    "AUC", 
    drop = TRUE
  ) %>% 
  {sprintf("(AUC = %s", gsub(" (", ", 95 % CI: ",., fixed = TRUE))}
}

@@

**RESULTS**

We found a parsimonious main effects model combining age, sex, American Society for Anesthesiologists (ASA) class, the presence of cancer, diseases of the central nervous system, kidney disease, and obesity. The model had good discrimination, both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`. This model was superior to traditional models based on the Charlson `r get_auc("Charlson")` and Elixhauser `r get_auc("Elixhauser")` co-morbidity indices. The model was well calibrated for predicted probabilities up to 5%.

**CONCLUSION:** We developed a personalized model to predict 90-day mortality after elective cemented THA based on easily available demographic data and the presence of four defined co-morbidities. This may facilitate individualized risk assessment prior to one of the most common surgical interventions, and a web-based calculator to aid clinical decision-making was published (`r shiny_url`).

@@

Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process. In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure. 90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.

@@

Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation). Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples. We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally. An AUC with 95 % confidence interval was calculated for the model as-is. Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales. An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.

@@

`r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)` % females) were included in the derivation cohort from SHAR (Figure \@ref(fig:flowchart) left panel). `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` of those patients died within 90 days and no one was censored before that. The unadjusted risk of 90-day mortality was therefore `r surv90d`. Further characteristics of the study population are presented in Table \@ref(tab:tab1). `r round(mean(df$CCI_index_quan_original != 0) * 100)` % and `r round(mean(df$ECI_index_sum_all  != 0) * 100)` % of the Swedish patients had at least one pre-surgery co-morbidity according to Charlson and Elixhauser. The proportion of patients with ASA class III was `r round(mean(df$P_ASA == 3) * 100)` %.

In addition, `r format(N_njr$tot, big.mark = ",")` patients were included for the external validation cohort from England and Wales (Figure \@ref(fig:flowchart) right panel). Their unadjusted risk of 90 day mortality was `r njr_surv`, which was higher compared to the derivation cohort from Sweden.

@@

The discriminative ability for the reduced model was not statistically significantly different when applied to the external validation cohort `r get_auc("NJR (Reduced model)")` compared to the internal validation `r get_auc("Reduced model")` (Figure \@ref(fig:rocs) right panel and Figure \@ref(fig:aucci)). Calibration of the re-calibrated model was slightly inferior compared to the internal calibration. Predicted probabilities between 0.5 % and 1.5 % were lower than observed proportions and the estimated 95 % confidence bands were wider. Over-all, calibration was still good however for predicted probabilities below 3 % and acceptable below 5 % (Figure \@ref(fig:calibration)).

@@

Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %). We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.

@@

In addition to data validity, the indications for choosing between different fixation techniques differ in Sweden compared to England and Wales. Cementation is the most commonly used fixation principle in Sweden but not so in England and Wales, where younger patients are more likely to be operated with cementless implants. To only include cemented THA therefore implies an over-representation of older and frailer patients from England and Wales compared to Sweden. Those patients might be more likely to die within 90 days, and this assumption would be in agreement with the calibration plot where estimated probabilities below 1.5 % underestimated the observed proportion of deaths in the external validation cohort from the NJR.

We investigated whether the effect of age on mortality was non-linear using restricted cubic splines. A similar approach might be relevant for BMI, which we only modelled as a linear main effect. A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel. Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

@@

@@

# Exclusive licence

We grant a worldwide licence to the publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution and convert or allow conversion into any format including without limitation audio, iii) create any other derivative work(s) based in whole or part on the on the Contribution, iv) to exploit all subsidiary rights to exploit all subsidiary rights that currently exist or as may exist in the future in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.

# Competing interests statement

All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare no financial relationships with any organisations that might have an interest in the submitted work in the previous three years, and no other relationships or activities that could appear to have influenced the submitted work.

# Transparency Statement

NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.

# Data sharing statement

Data is available upon request to the principal investigator.
